---
title: "Detecting hidden heterogeneity in single cell RNA-Seq data"
author: "Donghyung Lee"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


The iasva package can be used to detect hidden heterogenity within bulk or single cell sequencing data. To illustrate how to use the iasva package for heterogenity detection, we use real-world single cell RNA sequencing (scRNA-Seq) data obtained from human pancreatic islet samples ([Lawlor et. al., 2016](http://genome.cshlp.org/content/early/2017/01/16/gr.212720.116)). This dataset is included in a R data package ("iasvaExamples") containing data examples for IA-SVA (https://github.com/dleelab/iasvaExamples). To install the package, follow the instruction provided in the GitHub page.

## Install packages
```{r install_packages, echo=TRUE, eval=FALSE}
#devtools
library(devtools)
#iasva
devtools::install_github("UcarLab/IA-SVA")
#iasvaExamples  
devtools::install_github("dleelab/iasvaExamples")
```

## Load packages
```{r load_packages, echo=TRUE, message=FALSE}
rm(list=ls())
library(irlba) # partial SVD, the augmented implicitly restarted Lanczos bidiagonalization algorithm
library(iasva)
library(iasvaExamples)
library(sva)
library(Rtsne)
library(pheatmap)
library(corrplot)
library(DescTools) #pcc i.e., Pearson's contingency coefficient
library(RColorBrewer)

color.vec <- brewer.pal(3, "Set1")
```

## Load the islet single cell RNA-Seq data
```{r load_data, echo=TRUE}
data("Lawlor_Islet_scRNAseq_Read_Counts")
data("Lawlor_Islet_scRNAseq_Annotations")
ls()
counts <- Lawlor_Islet_scRNAseq_Read_Counts
anns <- Lawlor_Islet_scRNAseq_Annotations
dim(anns)
dim(counts)

summary(anns)
ContCoef(table(anns$Gender, anns$Cell_Type))
ContCoef(table(anns$Phenotype, anns$Cell_Type))
ContCoef(table(anns$Race, anns$Cell_Type))
ContCoef(table(anns$Patient_ID, anns$Cell_Type))
ContCoef(table(anns$Batch, anns$Cell_Type))
```

The annotations describing the islet samples and experimental settings are stored in "anns" and the read counts information is stored in "counts". 

## Extract alpha cells (GCG expressed cells) from non-diabetics 
To illustrate how IA-SVA can be used to detect hidden heterogeneity within a homogenous cell population (i.e., alpha cells), we use read counts of alpha cells from healthy (non-diabetic) subjects (n = 101).   
```{r alpha_cells, echo=TRUE, results='asis'}
counts <- counts[, (anns$Phenotype!="Non-Diabetic")&(anns$Cell_Type=="GCG")] 
anns <- subset(anns, (Phenotype!="Non-Diabetic")&(Cell_Type=="GCG"))
dim(counts)
dim(anns)

anns <- droplevels(anns)

prop.zeros <- sum(counts==0)/length(counts)
prop.zeros

# filter out genes that are sparsely and lowly expressed
filter = apply(counts, 1, function(x) length(x[x>5])>=3)

counts = counts[filter,]
dim(counts)

prop.zeros <- sum(counts==0)/length(counts)
prop.zeros
```

## Calculate geometric library size, i.e., library size of log-transfromed read counts 
It is well known that the geometric library size (i.e., library size of log-transfromed read counts) or proportion of expressed genes in each cell explains a very large portion of variability of scRNA-Seq data ([Hicks et. al. 2015 BioRxiv](http://biorxiv.org/content/early/2015/08/25/025528), [McDavid et. al. 2016 Nature Biotechnology](http://www.nature.com/nbt/journal/v34/n6/full/nbt.3498.html)). Frequently, the first principal component of log-transformed scRNA-Seq read counts is highly correlated with the geometric library size (r ~ 0.9). Here, we calculate the geometric library size vector, which will be used as a known factor in the IA-SVA algorithm.
```{r geno_lib_size, echo=TRUE, fig.width=7, fig.height=4}
Geo_Lib_Size <- colSums(log(counts+1))
barplot(Geo_Lib_Size, xlab="Cell", ylab="Geometric Lib Size", las=2)
lcounts <- log(counts + 1)

# PC1 and Geometric library size correlation
pc1 = irlba(lcounts - rowMeans(lcounts), 1)$v[,1] ## partial SVD
cor(Geo_Lib_Size, pc1)
```

## Run IA-SVA
Here, we run IA-SVA using Patient_ID and Geo_Lib_Size as known factors and identify five hidden factors. 
SVs are plotted in a pairwise fashion to uncover which SVs can seperate cell types. 
```{r run_iasva, echo=TRUE, fig.width= 7, fig.height=6}
set.seed(454353)
Patient_ID <- anns$Patient_ID
mod <- model.matrix(~Patient_ID+Geo_Lib_Size)
iasva.res<- iasva(t(counts), mod[,-1],verbose=FALSE, permute=FALSE, num.sv=5) ##irlba
iasva.sv <- iasva.res$sv

plot(iasva.sv[,1], iasva.sv[,2], xlab="SV1", ylab="SV2")

Cell_Type <- as.factor(iasva.sv[,2] > -0.2) 
levels(Cell_Type)=c("Cell1","Cell2")
table(Cell_Type)

# We identified 6 outlier cells based on SV2 that are marked in red
pairs(iasva.sv, main="IA-SVA", pch=21, col=color.vec[Cell_Type], bg=color.vec[Cell_Type], oma=c(4,4,6,12)) #4,4,6,12
legend("right", levels(Cell_Type), fill=color.vec, bty="n")

plot(iasva.sv[,1:2], main="IA-SVA", pch=21, xlab="SV1", ylab="SV2", col=color.vec[Cell_Type], bg=color.vec[Cell_Type])

cor(Geo_Lib_Size, iasva.sv[,2])

corrplot(cor(iasva.sv))
```

As shown in the above figure, SV2 clearly separates alpha cells into two groups: 6 outlier cells (marked in red) and the rest of the alpha cells (marked in green). SV3 and SV4 also capture outlier cells. However, we will focus on SV2 in the rest of the analyses.


## Find marker genes for the detected heterogeneity (SV2).
Here, using the find.markers() function we find marker genes (n=105 genes) that are significantly associated with SV2 (multiple testing adjusted p-value < 0.05, default significance cutoff, and R-squared value > 0.3, default R-squared cutoff).  
```{r find_markers, echo=TRUE, fig.width=7, fig.height=14}
marker.counts <- find.markers(t(counts), as.matrix(iasva.sv[,2]))
nrow(marker.counts)
rownames(marker.counts)

anno.col <- data.frame(Cell_Type=Cell_Type)
rownames(anno.col) <- colnames(marker.counts)
head(anno.col)

pheatmap(log(marker.counts+1), show_colnames =FALSE, clustering_method = "ward.D2",cutree_cols = 2,annotation_col = anno.col)
```



## Run tSNE to detect the hidden heterogeneity.
For comparison purposes, we applied tSNE on read counts of all genes to identify the hidden heterogeneity. We used the Rtsne R package with default settings.

```{r run_tsne, echo=TRUE, fig.width=7, fig.height=7}
set.seed(323542534)
tsne.res <- Rtsne(t(lcounts), dims = 2)

plot(tsne.res$Y, main="tSNE", xlab="tSNE Dim1", ylab="tSNE Dim2", pch=21, col=color.vec[Cell_Type], bg=color.vec[Cell_Type], oma=c(4,4,6,12))
legend("bottomright", levels(Cell_Type), fill=color.vec, bty="n")
```

As shown above, tSNE fails to detect the outlier cells that are identified by IA-SVA when all genes are used. Same color coding is used as above. 

## Run tSNE post IA-SVA analyses, i.e., run tSNE on marker genes associated with SV2 as detected by IA-SVA.
Here, we apply tSNE on the marker genes for SV2 obtained from IA-SVA
```{r run_tsne_post_iasva, echo=TRUE, fig.width=7, fig.height=7}
set.seed(345233)
tsne.res <- Rtsne(unique(t(log(marker.counts+1))), dims = 2)

plot(tsne.res$Y, main="tSNE post IA-SVA", xlab="tSNE Dim1", ylab="tSNE Dim2", pch=21, col=color.vec[Cell_Type], bg=color.vec[Cell_Type], oma=c(4,4,6,12))
legend("bottomright", levels(Cell_Type), fill=color.vec, bty="n")
```

tSNE using SV2 marker genes better seperate these ourlier cells. This analyses suggest that gene selection using IA-SVA combined
with tSNE analyses can be a powerful way to detect rare cells introducing variability in the single cell gene expression data. 

## Run principal component analysis (PCA) to detect the hidden heterogeneity (SV2).
Here, we use PCA to detect the hidden heterogeneity (SV2) detected by IA-SVA. 
```{r run_pca, echo=TRUE, fig.width=7, fig.height=6}
pca.res = irlba(lcounts - rowMeans(lcounts), 5)$v ## partial SVD

pairs(pca.res, main="PCA", pch=21, col=color.vec[Cell_Type], bg=color.vec[Cell_Type], oma=c(4,4,6,12)) #4,4,6,12
legend("right", levels(Cell_Type), fill=color.vec, bty="n")

plot(pca.res[,2:3], main="PCA", xlab="PC2", ylab="PC3", pch=21, col=color.vec[Cell_Type], bg=color.vec[Cell_Type], oma=c(4,4,6,12))
legend("bottomright", levels(Cell_Type), fill=color.vec, bty="n")
```

PC3 somewhat captures the six outlier cells, however this seperation is not as clear as the IA-SVA results.

## Run surrogate variable analysis (SVA) to detect the hidden heterogeneity (SV2).
Here, for comparison purposes we use SVA (using thre SVs) to detect the hidden heterogeneity in our example data. 
```{r run_sva, echo=TRUE, fig.width=7, fig.height=6}
mod1 <- model.matrix(~Patient_ID+Geo_Lib_Size)
mod0 <- cbind(mod1[,1])

sva.res = svaseq(counts,mod1,mod0, n.sv=5)$sv

pairs(sva.res, main="SVA", pch=21, col=color.vec[Cell_Type], bg=color.vec[Cell_Type], oma=c(4,4,6,12)) #4,4,6,12
legend("right", levels(Cell_Type), fill=color.vec, bty="n")

plot(sva.res[,1:2], main="SVA", xlab="SV1", ylab="SV2", pch=21, col=color.vec[Cell_Type], bg=color.vec[Cell_Type])
```

SV2 is associated with the six outlier samples, however the seperation of these cells is not as clear as the IA-SVA results.


## Correlation between SV2 and the geometric library size
```{r SV2_geometric_lib_size, echo=TRUE}
cor(Geo_Lib_Size, iasva.sv[,2])
```


## Session Info
```{r session_info, echo=TRUE}
sessionInfo()
```
