---
title: "IA-SVA based feature selection improves the performance of clustering algorithms [2]"
author: "Donghyung Lee"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


The IA-SVA based feature selection can improve the performance of clustering algorithms (e.g., tSNE or hierarchical clustering). To illustrate this, we used single cell RNA sequencing (scRNA-Seq) data obtained from human pancreatic islet samples ([Lawlor et. al., 2016](http://genome.cshlp.org/content/early/2017/01/16/gr.212720.116)). This dataset is included in an R data package ("iasvaExamples") containing data examples for IA-SVA (https://github.com/dleelab/iasvaExamples). To install the 'iasvaExamples' package, follow the instructions provided in the GitHub page.

## Load packages
```{r load_packages, echo=TRUE}
rm(list=ls())
library(iasva)
library(iasvaExamples)
library(Rtsne)
library(pheatmap)
```

## Load the islet single cell RNA-Seq data (638 genes, >26K genes)
```{r load_data, echo=TRUE}
data("Lawlor_Islet_scRNAseq_Read_Counts")
data("Lawlor_Islet_scRNAseq_Annotations")
ls()
counts <- Lawlor_Islet_scRNAseq_Read_Counts
anns <- Lawlor_Islet_scRNAseq_Annotations
dim(anns)
dim(counts)
# set seed for the random number generator.
set.seed(344532456)
```

The annotations describing the islet samples and experimental settings are stored in "anns" and the read counts information is stored in "counts". 

## Filter out low-expressed genes dataset reduced to 18K genes and 617 cells
```{r alpha_cells, echo=TRUE, results='asis'}
counts <- counts[, (anns$Cell_Type!="none")] 
anns <- subset(anns, (Cell_Type!="none"))
dim(counts)
dim(anns)

anns <- droplevels(anns)

prop.zeros <- sum(counts==0)/length(counts)
prop.zeros

filter = apply(counts, 1, function(x) length(x[x>5])>=3)

counts = counts[filter,]
dim(counts)

prop.zeros <- sum(counts==0)/length(counts)
prop.zeros

Patient_ID <- anns$Patient_ID
Cell_Type <- anns$Cell_Type
Phenotype <- anns$Phenotype
Batch <- anns$Batch
```

## Calculate geometric library size, i.e., library size of log-transfromed read counts 
It is well known that the geometric library size (i.e., library size of log-transfromed read counts) or proportion of expressed genes in each cell explains a very large portion of variability of scRNA-Seq data ([Hicks et. al. 2015 BioRxiv](http://biorxiv.org/content/early/2015/08/25/025528), [McDavid et. al. 2016 Nature Biotechnology](http://www.nature.com/nbt/journal/v34/n6/full/nbt.3498.html)). Frequently, the first principal component of log-transformed scRNA-Seq read counts is highly correlated with the geometric library size (r > 0.9). Here, we calculate the geometric library size to be used as a known factor in the IA-SVA analyses.
```{r geno_lib_size, echo=TRUE, fig.width=6, fig.height=4}
Geo_Lib_Size <- colSums(log(counts+1))
barplot(Geo_Lib_Size, xlab="Cell", las =2)
lcounts <- log(counts + 1)
pca.res = svd(lcounts - rowMeans(lcounts))$v
cor(Geo_Lib_Size, pca.res[,1])
```

## Run tSNE to cluster islet cells.
For comparison purposes,we applied tSNE on read counts of all genes. We used the Rtsne R package with default settings.
Genes are color coded wrt their expression of marker genes. 
INS (beta), GCG(alpha), SST(delta), PPY(PP/gamma), COL1A1 (stellate), PRSS1 (acinar), KRT19 (ductal) cells.   

```{r run_tsne, echo=TRUE, fig.width=6, fig.height=6}
tsne.res <- Rtsne(t(lcounts), dims = 2)

plot(tsne.res$Y, main="tSNE", pch=21, bg=c("red","green3","blue","grey","orange","black","purple")[Cell_Type], oma=c(4,4,6,12), cex=1.5)
legend("topleft", levels(Cell_Type), fill=c("red", "green3", "blue","grey","orange","black","purple"), bty="n")
```

## Run IA-SVA
Here, we run IA-SVA using Phenotype, Patient_ID, Batch and Geo_Lib_Size as known factors and identify 6 hidden factors. 
```{r run_iasva, echo=TRUE, fig.width= 6, fig.height=5}
mod <- model.matrix(~Phenotype+Patient_ID+Batch+Geo_Lib_Size)
iasva.res<- iasva(t(counts), mod[,-1],verbose=FALSE, permute=FALSE, num.sv=6)
iasva.sv <- iasva.res$sv

pairs(iasva.sv, main="tSNE", pch=21, bg=c("red","green3","blue","grey","orange","black","purple")[Cell_Type], oma=c(4,4,6,12))
legend(0.82, 0.65, levels(Cell_Type), fill=c("red", "green3", "blue","grey","orange","black","purple"), bty="n")
```

As shown in the above figure, SV1, SV3, SV4 and SV6 separate cells.


## Find marker genes for SV1, SV3, SV4 and SV6.
Here, we find marker genes significantly associated with SV1, SV3, SV4 and SV6 (multiple testing adjusted p-value < 0.05, default significance cutoff) and having a high R-squared value (R-squared > 0.4) using the find.markers() function.
This generated 195 marker genes, that also includes well-established markers INS, GCG, COL1A1, PRSS1
```{r find_markers, echo=TRUE, fig.width=6, fig.height=5}
marker.counts <- find.markers(t(counts), as.matrix(iasva.sv[,c(1,3,4,6)]),  rsq.cutoff = 0.4)
nrow(marker.counts)

anno.col <- data.frame(Cell_Type=Cell_Type)
rownames(anno.col) <- colnames(marker.counts)
head(anno.col)
pheatmap(log(marker.counts+1), show_colnames =FALSE, show_rownames = TRUE, clustering_method = "ward.D2",cutree_cols = 7,annotation_col = anno.col)
```

## Run tSNE post IA-SVA, i.e., run tSNE on 195 marker genes obtained from IA-SVA.
Here, we apply tSNE on the marker genes for SV2 obtained from IA-SVA
```{r run_tsne_post_iasva, echo=TRUE, fig.width=6, fig.height=6}
set.seed(75458456)
tsne.res <- Rtsne(unique(t(log(marker.counts+1))), dims = 2)
plot(tsne.res$Y, main="tSNE", pch=21, bg=c("red","green3","blue","grey","orange","black","purple")[Cell_Type], oma=c(4,4,6,12), cex=1.5)
legend("topright", levels(Cell_Type), fill=c("red", "green3", "blue","grey","orange","black","purple"), bty="n")
```

tSNE performed on marker genes selected via IA-SVA performs better than original tSNE analyses using all genes. 
This example again reiterates the importance of gene selection using IA-SVA for effective clustering of single-cell datasets. 
## Session Info
```{r session_info, echo=TRUE}
sessionInfo()
```
